The late 1800's debate over the age of the Earth: an early attempt of interdisciplinary science

### 2023-08-16 ms.tex
sec 3 and 4

The purpose of this paper is to attempt to shed light on why interdisciplinary science failed during the late 1800s debate over the age of the Earth. Furthermore, I use this historical failure to argue how interdisciplinary science might be best carried out in modern research. To achieve these goals, I first summarize the late 1800s discourse over Earth's age (Section \ref{sec:2}). Next, in Section \ref{sec:3}, I provide possible reasons for why this early attempt of interdisciplinary science was unsuccessful, and what lessons might be drawn from history to improve modern interdisciplinary work. Finally, in Section \ref{sec:4}, I summarize and conclude.

@@

\section{A reason why late 1800s interdisciplinary science was unsuccessful and implications for modern research} \label{sec:3}

@@

There are likely many reasons why interdisciplinary science failed. I believe one important reason was that, at the time, different disciplines were culturally tied to specific kinds of scientific evidence. Thanks to Lyell's paradigm of uniformitarianism, the late 1800s geologist took a qualitative approach, studying the Earth by solely examining piles of strata. This method for understanding the world was part of what it meant to be a geologist. In contrast, a physicist was defined by a mathematical or quantitative approach. Given this cultural division between the disciplines, no researcher was in a position to consider all evidence on equal footing.

\section{Conclusions} \label{sec:4}

In the 1860s, Lord Kelvin used his understanding of physics and thermodynamics to estimate the age of the Earth. He brought these calculations to the attention of geologists, arguing against the uniformitarian view of an indefinitely old Earth and troubling Darwin, who wanted vast periods of history for his theory of natural selection. The ensuing debate, involving Lyell's geology, Darwin's biology, and Kelvin's physics, is an early attempt of interdisciplinary science. However, Kelvin's quantitative arguments for a young Earth were not entertained or understood by large portion of the geologic community for about a decade. Here, I argue that one reason this interdisciplinary science failed to proceed smoothly is because, at the time, there were big cultural divisions between the quantitative physical sciences and the more qualitative geologic and biological sciences. Today, some of the big unsolved problems in science will likely require collaborations between many different disciplines. To solve these big problems, perhaps we can learn from the failed 1860s debate over the age of the Earth, by striving to break down cultural and tribal barriers between different fields of study.

>>>>

The purpose of this paper is to attempt to shed light on why interdisciplinary science failed during the late 1800s debate over the age of the Earth. Furthermore, I compare this historical failure to modern interdisciplinary science. To achieve these goals, I first summarize the late 1800s discourse over Earth's age (Section \ref{sec:2}). Next, in Section \ref{sec:3}, I provide possible reasons for why this early attempt of interdisciplinary science was unsuccessful, and juxtapose this history to modern interdisciplinary work. Finally, in Section \ref{sec:4}, I summarize and conclude.

@@

\section{The reasons why late 1800s interdisciplinary science was unsuccessful and a comparison to modern research} \label{sec:3}

@@

There are likely many reasons why interdisciplinary science failed. I believe one important reason was that, at the time, different disciplines were culturally tied to specific kinds of scientific evidence. Thanks to Lyell's paradigm of uniformitarianism, the late 1800s geologist took a qualitative approach, studying the Earth by solely examining piles of strata. This method for understanding the world was part of what it meant to be a geologist. In other words, a commitment to the stratigraphic record was part of a geologist's identity. In contrast, a physicist was defined and identified by a mathematical or quantitative approach. Given this cultural division between the disciplines, no researcher was in a position to consider all evidence on equal footing.

In modern science, disciplines are not as entangled with specific methods or evidence relevant to understanding the world. Today, geologists will often attempt use all the evidence at their disposal to solve the question at hand regardless of whether information is qualitative or quantitative.

However, while a modern scientist may be inclined to consider all available information, they can only fully evaluate evidence within their specialization. All modern researchers need to specialize in certain information-gathering techniques. For example, a worker could be an expert in clumped isotope mass spectroscopy, finite element geodynamic modeling, or full-waveform seismic tomography. All these different methods may be relevant to the same unsolved geologic question, but any single person can only hope to fully understand one (or maybe two) of these approaches.

The need for specialization today, places researchers in similar positions of ignorance as Kelvin, Huxley and Darwin in their debate over the age of the Earth. Kelvin was unable to evaluate qualitative stratigraphic evidence, and geologists could not understand Kelvin's mathematics. In the same way, the modern geologist attempting to address an interdisciplinary problem like the origin of life, will not be able to accurately judge evidence from, for example, the organic chemistry literature.

A further problem is that specialization incentivizes researchers to emphasize or over-weight evidence from their domain of expertise. Consider the putative discovery of methane in Mars' atmosphere by the Curiosity rover mass spectrometer \citep{Webster_2015}. Methane on Mars is potentially a very important discovery, because on Earth, methane gas is associated with life. But the problem with the Curiosity observations is that the Mars Trace Gas Orbiter could not duplicate the observation \citep{Korablev_2019}. Also, methane at the concentrations suggested by Curiosity is not compatible with our understanding of methane photochemistry \citep{Lefevre_2009}. Despite this contradicting evidence researchers specialized in the Curiosity rover mass spectrometer are incentivized to champion the detection, because if there is indeed methane on Mars created by life, then the mass spectroscopists will be credited for the revolutionary discovery.

In the same vein, Kelvin was incentivized to highlight his calculations for Earth's age, because if they were accurate, then he would be responsible for an important new understanding of the world. On the other hand, geologists were motivated to ignore Kelvin, because they were invested in studying Earth through the stratigraphic record.

In summary, I suggest that one reason why interdisciplinary science failed to proceed smoothly during the late 1800s debate over the age of the Earth was because each discipline was culturally obligated to consider only some of the evidence. At the time, geology was defined by a qualitative interrogation of the stratigraphic record, and physics was instead a mathematical undertaking. The need to specialize in modern research puts scientists in a similar position to Kelvin, Darwin, Huxley and their contemporaries. Specialization makes it challenging for scientists today to understand all the evidence relevant to interdisciplinary problems, and also incentivizes workers to over-weight information generated by their sub-field of study.

This is not an argument against specialization - it is a necessary part of scientific advancement. Instead, I believe we should be aware that specialization may cause interdisciplinary science to be more challenging.

\section{Summary and conclusions} \label{sec:4}

In the 1860s, Lord Kelvin used his understanding of physics and thermodynamics to estimate the age of the Earth. He brought these calculations to the attention of geologists, arguing against the uniformitarian view of an indefinitely old Earth and troubling Darwin, who wanted vast periods of history for his theory of natural selection. The following debate, involving Lyell's geology, Darwin's biology, and Kelvin's physics, is an early attempt of interdisciplinary science. However, Kelvin's quantitative arguments for a young Earth were not entertained or understood by large portion of the geologic community for about a decade. Here, I argue that one reason this interdisciplinary science failed to proceed smoothly is because, at the time, there were big cultural divisions between physics and biology/geology that compelled researchers to evaluate only some of the evidence. Physics was a exclusively a quantitative endeavor, while under Lyell's paradigm of uniformitarianism, geology was a qualitative investigation of the stratigraphy.

I also suggest that specialization in certain information-gathering techniques can put modern-day researchers in a similar position to those involved in the debate over Earth's age in the late 1800s. Specialization can incentivizes a researcher to prefer their evidence over others, and also ill-prepare a worker to understand all information relevant to an interdisciplinary problem.
